\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-18 paper.tex

Tremendous progress has been made in recent years on the long-standing
problem of ARG inference, there is now a range of practically 
applicable methods available.
Methods targetting large-scale datasets tend to simplify
the inference problem by
making a single, deterministic 
best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast,
and it is therefore not clear that generating many guesses
when sample sizes are large will achieve much in terms of capturing uncertainty.
An alternative approach to 
is to incorporate uncertainty encountered during inference into the returned ARG.
The gARG encoding described here enables particular kinds of uncertainty
to be incorporated directly into the topology:
nodes that have more than two children (polytomies)
represent uncertainty over the ordering of coalescence events 
(Appendix~\ref{sec-cell-lineages-and-args}), 
and those have have more than two parents 
represent uncertainty over the ordering of multiple recombination events
(Appendix~\ref{sec-ARG-simplification}). 
Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be utilised in downstream applications
is an open question.

>>>>

Tremendous progress has been made in recent years on the long-standing
problem of ARG inference, there is now a range of practically 
applicable methods available.
Methods targeting large-scale datasets tend to simplify
the inference problem by
making a single, deterministic 
best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast,
and it is therefore not clear that generating many guesses
when sample sizes are large will achieve much in terms of capturing uncertainty.
An alternative approach to 
is to incorporate uncertainty encountered during inference into the returned ARG.
The gARG encoding described here enables particular kinds of uncertainty
to be incorporated directly into the topology:
nodes that have more than two children (polytomies)
represent uncertainty over the ordering of coalescence events 
(Appendix~\ref{sec-cell-lineages-and-args}), 
and those have have more than two parents 
represent uncertainty over the ordering of multiple recombination events
(Appendix~\ref{sec-ARG-simplification}). 
Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be utilised in downstream applications
is an open question.
